### 2015-12-05 paper/1-introduction.tex
Last? changes. Increase memory interconnect ram profile.

\subsection{Solution} 
%TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
The usage of the Direct connection algorithm makes reusing adevs kernel hard. Devs-ex-machina (dxex) is based on PythonPDEVS \cite{PythonPDEVS}(PyPDEVS) but differs in design and implementation.
The core aim of the project is to offer a deterministic simulation kernel where the simulation modeller is shielded as much as possible from the kernel implementation, without sacrificing performance. As in PyPDEVS, a model need be written only once for use in the different simulation kernels (with the exception of a non-trivial lookahead).\\
The tracing framework from PyPDEVS was ported to allow optional verification of simulations. 
\subsection{Time}
The DEVS formalisms have $\mathbb{R}$ as time base, but any implementation has to decide on an enumerable representation of time. In principle any type with well defined operators can be used as template parameter, but from a performance point of view a type fitting in a machine word offers obvious advantages. In practice, dxex defaults to word sized unsigned integer.
Although this significantly reduces the possible range of virtual time, it avoids approximation errors. Furthermore, the notion of $\epsilon$ as a minimum distance between two distinct time points need not to be established. This task is non-trivial for floating point approximations.\\
Our time representation is extended with a causality field with a range equal or greater than the maximum nr of models in a single kernels ($2^{48}$ by default). If A and B are imminent at time t then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$. This allows a deterministic ordering of concurrent events. % execution not so much as tracing

>>>>

\subsection{Solution} 
%TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
The usage of the Direct connection algorithm makes reusing adevs kernel hard. Devs-ex-machina (dxex) is based on PythonPDEVS \cite{PythonPDEVS}(PyPDEVS) but differs in design and implementation.
The core aim of the project is to offer a deterministic simulation kernel where the simulation modeller is shielded as much as possible from the kernel implementation, without sacrificing performance. As in PyPDEVS, a model need be written only once for use in the different simulation kernels (with the exception of a non-trivial lookahead).\\
The tracing framework from PyPDEVS was ported to allow optional verification of simulations. 
\subsection{Time}
The DEVS formalisms have $\mathbb{R}$ as time base, but any implementation has to decide on an enumerable representation of time. In principle any type with well defined operators can be used as template parameter, but from a performance point of view a type fitting in a machine word offers obvious advantages. In practice, dxex defaults to word sized unsigned integer.
Although this significantly reduces the possible range of virtual time, it avoids approximation errors. The notion of $\epsilon$ is trivial for integer, unlike floating point.\\
Our time representation is extended with a causality field with a range equal or greater than the maximum nr of models in a single kernel ($2^{48}$ by default). If A and B are imminent at time t then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$. This allows a deterministic ordering of concurrent events. % execution not so much as tracing

### 2015-12-05 paper/4-performance.tex
Last? changes. Increase memory interconnect ram profile.

%TODO say that you will now prove your claims about the aproach having nearly no impact
%TODO and potentially even claim that you are faster than adevs for some situations, due to the algorithms from PythonPDEVS and further features
%TODO briefly give some hardware info on the machine (CPU, OS, RAM), used software (version of adevs, GCC, ...), and about number of samples/variance
\subsection{Sequential Simulation}
%TODO we start with sequential simulation results, because this is still base case
%TODO compare CPU time with adevs for some benchmarks
%TODO and explain why there is such a difference
%TODO (spend some time here, as this is also important)
\subsubsection{Preamble}
The benchmark models are identical in functionality (between adevs/dxex versions). Dxex uses integer timestamps, adevs uses floating point. Tests show there is little difference performance wise (both types fit in a machine word)\\
Benchmarks ran on a i7-2670QM, 16GiB Ram, Arch Linux (4.2.5 kernel), G++ 5.2. Interpretation of the benchmark results is done using call-graphs generated by perf, a selection of those call-graphs can be found in the code repository. For the runtime plots perf is used in combination with R.\\
Note that a fixed timeadvance (e.g. devstone) is set at 100.
\subsubsection{Devstone}
The Devstone \cite{DEVStone} benchmark is highly hierarchical, using Direct connection the dxex kernels can exploit this whereas adevs needs to walk the coupled model's structure to pass events.\\
\begin{figure}[h]
	\includegraphics[width=.5\textwidth]{fig/fig1.eps}
	\label{fig1.eps}
\end{figure}

@@

\subsection{Parallel Simulation}
%TODO explain that we will now compare different synchronization protocols
%TODO compare different synchronization protocols with adevs' conservative, showing that we are faster (or at least, not that much slower) than adevs
\subsubsection{Devstone}
The flattened models are allocated to kernels by giving each kernel a distinct section of the chain, resulting in a low ratio of inter-kernel to intra-kernel messages. For optimistic, this can cause more reverts since the kernels will start to drift faster as the model-count increases. Furthermore, optimistic is quite sensitive to an increase in kernels, since the delay before a revert propagates increases. Of note as well is the warm-up time this benchmark requires, for $n=d\times w$ models, it takes $\textup{timeadvance}()*(n-1)$ transitions to activate the last model in the chain. For parallel this can be reduced to $\textup{timeadvance}()*(n\frac{kernels-1}{kernels} -1)$ before the last kernel becomes active.\\
\begin{figure}[ltbh]
	\includegraphics[width=.5\textwidth]{fig/fig2.eps}
	\label{fig2.eps}
\end{figure}
\subsubsection{PHold}
%Phold
In Phold allocation is specified in the benchmark itself. Each kernel manages a single node with a constant set of sub-nodes. The parameter R (here 10) determines the percentage of remote destination models.\\
The dynamic dependency graph is a very sparse version of the static dependency graph, penalizing conservative. Lookahead is $\epsilon$, so conservative spends most of its time crawling in steps of $\epsilon$. Since the dependency graph between kernels is a complete graph, this is not a simulation that scales in our implementation. For N kernels, each kernel has to query the null-time of N-1 kernels, resulting in O($N^2$) polling behaviour. In a non-cyclic simulation with a non-trivial lookahead (e.g. devstone), that choice however does pay off. Adevs's lesser performance is due in part to their lookahead management, which after profiling shows to spend a non-trivial amount of time in exception handling code. 
Optimistic suffers little from the above problems, however due to the high interconnectivity a cascading revert is still possible. With the percentage of remotes = 100 phold reflects interconnect in behaviour, which is why the R value is not dimensioned here. A revert is very expensive in PHold due to our usage of C++11's random nr generators. The cost of a revert is dominated by the recalculation of destination models, not in allocating/deallocating states/messages. Again, this could be significantly reduced using lazy-cancellation. Once a revert happens the drift between kernels increases fast, increasing the likelihood of more reverts. Despite all this, optimistic can for low R values quickly exploit uncertainty that slows down conservative in this benchmark.

@@

Adevs passes messages by value (in container by reference), we pass by pointer. The runtime effects of this choice are already demonstrated in the interconnect benchmark, so in this section, we measure memory usage in number of allocated pages combining text, stack and heap memory for the program profiled. For the OS and/or user this is the actual memory footprint of the application. It is important to note that, especially in the case of optimistic, not all this memory is always in use by the kernels. During simulation, the pools will generally not return deallocated memory to the OS, but keep it for later reuse.

@@

Additionally, the differences in LP virtual times are far larger compared to conservative time synchronization. All these factors explain the heavier memory usage. Devstone (flattened) is allocated in a chain. Leafs in the dependency graph will therefore do a lot of unnecessary simulation before having a revert, leading to severe memory pressure. Unlike conservative and sequential execution, memory usage in optimistic varies greatly depending on scheduling of kernel threads and drifting between kernels. 

@@

With lookahead = $\epsilon$ this then leads to a high frequency of polling on shared null-times, which are used to determine lbts and thus garbage collection.

@@

Adevs' conservative fails to complete the benchmark under valgrind even with a significantly reduced load.\\

@@

	\caption{Interconnect w 10 t5e5, unit MiB, 2 kernels (if parallel)}

@@

		39 & 39 & 35 & 43 & 164 \\ \hline

>>>>

%TODO say that you will now prove your claims about the aproach having nearly no impact
%TODO and potentially even claim that you are faster than adevs for some situations, due to the algorithms from PythonPDEVS and further features
%TODO briefly give some hardware info on the machine (CPU, OS, RAM), used software (version of adevs, GCC, ...), and about number of samples/variance
\subsection{Sequential Simulation}
%TODO we start with sequential simulation results, because this is still base case
%TODO compare CPU time with adevs for some benchmarks
%TODO and explain why there is such a difference
%TODO (spend some time here, as this is also important)
\subsubsection{Preamble}
The benchmark models are identical in functionality (between adevs/dxex versions). Dxex uses integer timestamps, adevs uses floating point. Tests show there is little difference performance-wise (both types fit in a machine word).\\ The benchmarks are compiled using O3 and link-time optimization, with all IO disabled. The benchmarks have testing variants where the exact output can be verified.
Benchmarks ran on a i7-2670QM, 16GiB Ram, Arch Linux (4.2.5 kernel), G++ 5.2. Interpretation of the benchmark results is done using call-graphs generated by perf, a selection of those call-graphs can be found in the code repository. For the runtime plots perf is used in combination with R.\\
Note that a fixed timeadvance (e.g. devstone) is set at 100.
\subsubsection{Devstone}
The Devstone \cite{DEVStone} benchmark is highly hierarchical, using Direct connection the dxex kernels can exploit this whereas adevs needs to walk the coupled model's structure to pass events.\\
\begin{figure}[h]
	\includegraphics[width=.5\textwidth]{fig/fig1.eps}
	\label{fig1.eps}
\end{figure}

@@

\subsection{Parallel Simulation}
%TODO explain that we will now compare different synchronization protocols
%TODO compare different synchronization protocols with adevs' conservative, showing that we are faster (or at least, not that much slower) than adevs
Whenever not specified, the benchmarks use 4 kernels for parallel simulation.
\subsubsection{Devstone}
The flattened models are allocated to kernels by giving each kernel a distinct section of the chain, resulting in a low ratio of inter-kernel to intra-kernel messages. For optimistic, this can cause more reverts since the kernels will start to drift faster as the model-count increases. Furthermore, optimistic is quite sensitive to an increase in kernels, since the delay before a revert propagates increases. Of note as well is the warm-up time this benchmark requires, for $n=d\times w$ models, it takes $\textup{timeadvance}()*(n-1)$ transitions to activate the last model in the chain. For parallel this can be reduced to $\textup{timeadvance}()*(n\frac{kernels-1}{kernels} -1)$ before the last kernel becomes active.\\
\begin{figure}[ltbh]
	\includegraphics[width=.5\textwidth]{fig/fig2.eps}
	\label{fig2.eps}
\end{figure}
\subsubsection{PHold}
%Phold
In Phold allocation is specified in the benchmark itself. Each kernel manages a single node with a constant set of sub-nodes. The parameter R (here 10) determines the percentage of remote destination models.\\
The dynamic dependency graph is a very sparse version of the static dependency graph, penalizing conservative. Lookahead is $\epsilon$, so conservative spends most of its time crawling in steps of $\epsilon$. Since the dependency graph between kernels is a complete graph, this is not a simulation that scales in our implementation. For N kernels, each kernel has to query the null-time of N-1 kernels, resulting in O($N^2$) polling behaviour. In a non-cyclic simulation with a non-trivial lookahead (e.g. devstone), that choice however does pay off. Adevs's lesser performance is due in part to their lookahead management, which after profiling shows to spend a non-trivial amount of time in exception handling code. 
Optimistic suffers little from the above problems, however due to the high interconnectivity a cascading revert is still possible. With the percentage of remotes = 100 phold reflects interconnect in behaviour, which is why the R value is not dimensioned here. A revert is very expensive in PHold due to our usage of C++11's random nr generators. The cost of a revert is dominated by the recalculation of destination models, not in allocating/deallocating states/messages. Once a revert happens the drift between kernels increases fast, increasing the likelihood of more reverts. Despite all this, optimistic can for low R values quickly exploit the uncertainty that slows down conservative in this benchmark.

@@

Adevs passes messages by value, we pass by pointer. The runtime effects of this choice are already demonstrated in the interconnect benchmark, so in this section, we measure memory usage in number of allocated pages combining text, stack and heap memory for the program profiled. For the OS and/or user this is the actual memory footprint of the application. It is important to note that, especially in the case of optimistic, not all this memory is always in use by the kernels. During simulation, the pools will generally not return deallocated memory to the OS, but keep it for later reuse.

@@

Additionally, the differences in LP virtual times are far larger compared to conservative time synchronization. All these factors explain the heavier memory usage. Devstone (flattened) is allocated in a chain. Leafs in the dependency graph will therefore do a lot of unnecessary simulation before having a revert, leading to increased memory pressure. Unlike conservative and sequential execution, memory usage in optimistic varies greatly depending on scheduling of kernel threads and drifting between kernels. 

@@

With lookahead $\epsilon$ this then leads to a high frequency of polling on shared null-times, which are used to determine lbts and thus garbage collection.

@@

Adevs' conservative fails to complete the benchmark under valgrind even with a significantly reduced load.

@@

	\caption{Interconnect w 20 t5e5, unit MiB, 2 kernels (if parallel)}

@@

		39 & 39 & 35 & 43 & 259 \\ \hline

### 2015-12-05 paper/6-conclusions.tex
Add memory usage phold and offer explaination for increased memory usage conservative.
Last? changes. Increase memory interconnect ram profile.

%TODO conclude the paper, saying that our new simulation tool offers conservative and optimistic synchronization, resulting in good performance, which is even better than adevs
Sequentially dxex outperforms adevs in hierarchical simulations while yielding to adevs in broadcast-like simulations. 
Parallel the results are more nuanced. The sensitivity to broadcast simulations is carried over in the parallel implementation, resulting in performance loss for both synchronization protocols. Optimistic can still achieve good performance if the volume of inter-kernel messages is low. Both  are fast in non-cyclic simulations, with conservative better suited whenever state-saving becomes expensive and optimistic when lookahead is not available.\\
Finally, the current optimistic implementation is ill-suited for memory constrained systems, especially in simulations where the frequency of reverts is high.
\subsection{Future work}
%TODO future work: more things that we might want to implement, and probably something about automatically switching between synchronization protocols at runtime (and reference my activity papers ;))
\subsubsection{Memory}
Our optimistic implementation can benefit from a faster GVT algorithm such as FujiMoto's \cite{Fujimoto:1997:CGV:268403.268404} or more recently \cite{Bauer:2005:SND:1069810.1070159}. 
With access to the memory subsystem in place (pools), optimistic could be constrained using per kernel quota preventing the simulation from exhausting memory. While less sensitive to memory pressure, the same approach could benefit conservative.
\subsubsection{Activity}
As shown in \cite{PythonPDEVS_ACTIMS} activity and allocation of models across kernels is a key aspect in achieving high performance in any parallel implementation. Allocating models so that there are no dependency cycles between their containing kernels is a first step, but not always possible. For optimistic one can use re-allocation to break (runtime dependency) cycles and/or perform load-balancing. If kernels are unevenly balanced they will begin to drift fast, causing increasingly more reverts. There is a limited framework implemented to track model activity for debugging purposes, this could be extended to enable the above strategies.
\subsubsection{Hybrid}
The optimistic implementation could use (null/eot/eit) from conservative to detect and/or reduce the cost of reverts without completely stalling on influencing kernels.
Conservative kernels could be extended with runtime information about influencing models, for example if one can guarantee a static dependency is not used for a fixed time-span, this can be removed for that period of (virtual) time. 
Ultimately the simulation could switch at runtime between protocols based on the information provided by activity tracking. This requires the above mentioned framework, and could only be done at timepoints where a GVT/LBTS time is agreed upon between the kernels. From the model point of view no changes are needed, although a non-trivial lookahead is obviously desired.

>>>>

%TODO conclude the paper, saying that our new simulation tool offers conservative and optimistic synchronization, resulting in good performance, which is even better than adevs
Sequentially dxex outperforms adevs in hierarchical simulations while yielding to adevs in broadcast-like simulations. 
Parallel the results are more nuanced. The sensitivity to broadcast simulations is carried over in the parallel implementation, resulting in performance loss for both synchronization protocols. Optimistic can still achieve good performance if the volume of inter-kernel messages is low. Both  are fast in non-cyclic simulations, with conservative better suited whenever state-saving becomes expensive and optimistic when lookahead is not available.\\
Finally, the current optimistic implementation is ill-suited for memory constrained systems, especially in simulations where the frequency of reverts is high.
\subsection{Future work}
%TODO future work: more things that we might want to implement, and probably something about automatically switching between synchronization protocols at runtime (and reference my activity papers ;))
\subsubsection{Memory}
Our optimistic implementation can benefit from a faster GVT algorithm such as FujiMoto's \cite{Fujimoto:1997:CGV:268403.268404}  or more recently Bauer et al\cite{Bauer:2005:SND:1069810.1070159}. 
With access to the memory subsystem already in place (pools), optimistic could be constrained using per kernel quota preventing the simulation from exhausting memory. While less sensitive to memory pressure, the same approach could benefit conservative.
\subsubsection{Activity}
As shown in \cite{PythonPDEVS_ACTIMS} activity and allocation of models across kernels is a key aspect in achieving high performance in any parallel implementation. Allocating models so that there are no dependency cycles between their containing kernels is a first step, but not always possible. For optimistic one can use re-allocation to break (runtime dependency) cycles and/or perform load-balancing. If kernels are unevenly balanced they will begin to drift fast, causing increasingly more reverts. There is a limited framework implemented to track model activity for debugging purposes, this could be extended to enable the above strategies.
\subsubsection{Hybrid}
The optimistic implementation could use (null/eot/eit) from conservative to detect and/or reduce the cost of reverts without completely stalling on influencing kernels.\\
Conservative's eot calculation can be enhanced by using the entire dependency graph between kernels (not models). The implementation could detect dependency cycles between kernels and, instead of waiting on dependent kernels, advance as a synced group with better scaling. Lbts calculation could leverage this as well to reduce memory usage, though this is not that stringent in conservative.
Ultimately the simulation could switch at runtime between protocols based on the information provided by activity tracking. This requires the above mentioned framework, and could only be done at timepoints where a GVT/LBTS time is agreed upon between the kernels. From the model point of view no changes are needed, although a non-trivial lookahead is obviously desired.
